Sources of Measurement Bias in Modern CPUs

### 2014-01-28 paper.tex
Remove section on LSD, and start cleanup lplot

\documentclass[a4paper,10pt,twocolumn,twoside]{article}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{filecontents}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{tikz}
\usepackage{listings}
\usepackage{authblk}
\usepackage{caption} % For margin top bottom

@@

@@

@@

\begin{lstlisting}[float=h, language=C, caption={Naive implementation of convolution. Highly sensitive to aliasing between input and output arrays.}, label={lst:conv}, frame=lines]
static float k[5]={0.1,0.25,0.3,0.25,0.1};
void conv(int n, float *in, float *out) {
    int i, j;
    for (i = 2; i < n-2; ++i) {
      out[i] = 0;
      for (j = 0; j < 5; ++j)
          out[i] += in[i-2+j] * k[j];
    }
}
\end{lstlisting}

@@

\section{The Loop Stream Detector}
Typical software spends most of the time executing the same instructions repeatedly in loops, where the same instructions are fetched, decoded and executed many times. 
The Loop Stream Detector (LSD) is a front-end hardware optimization that is able to detect small software loops.
A cache of decoded instructions is kept and analyzed for cyclic chains in the instruction stream.
When a loop is detected, can be fetched directly from a queue of already decoded micro operations.
This enables some instruction decode logic in the processor front end to be temporarily disabled.
The LSD is a power saving optimization, but can also improve performance in cases where the decode pipeline is a bottleneck.
Utilization of the Loop Stream Detector can be measured by the following performance counter.
\begin{description}
  \item{LSD.UOPS\marginpar{\vspace{0.0pt}\footnotesize\begin{tabular}{@{} l @{\hskip 2pt} l }Event & Mask \\ 0xA8 & 0x01\end{tabular}}} \hfill \\
  Counts the number of micro-operations delivered by the Loop Stream Detector.
\end{description}

The LSD has been suggested in previous work as a potential source of measurement bias [Mytkowicz:2009:WrongData].
The Loop Stream Detector is dependent on code layout, which can change with external factors such as variations in link ordering. 
Interacting with the alignment and position of instructions in memory, the LSD can cause performance differences for programs with otherwise \emph{identical} instruction sequences.

Some effort has been made to model properties of the LSD on previous Intel architectures, as an aid in low-level optimization. 
Hundt et al.~\cite{Hundt:2011:MAO} looks at assembly rewrite rules for optimizing agains the Core 2 LSD.
For loops to be recognized, they have to match a set of restrictions imposed by the hardware implementation that tries to detect them. 
For instance, because the LSD cache is of fixed size there will be a maximum number of instructions a loop can contain for it to be recognized.
The capabilities of the LSD has improved with newer micro-architectures.
Some important limitations for Haswell's LSD are listed below.
\begin{itemize}
  \item Up to 12 chunk fetches of 32 B instruction memory.\footnote{The optimization manual actually states 8 chunk fetches for HTT enabled, and 11 for HTT disabled. Our experiments with a minimal test case shows that the LSD is active for loops spanning up to 12 chunks -- regardless of hyper threading.}
  \item Up to 28 micro-operations for HTT enabled, 56 for HTT disabled.
  \item Can not contain CALL or RET instructions
  \item Stack operations must match, i.e push must be matched with a pop.
\end{itemize}

\subsection{Bias from Link Order}
Different link orders affect the relative position of code and static data within the final compiled binary file.
More importantly, \emph{alignment} of code can be affected by link order.
A loop that was aligned to 32 byte for one ordering might end up at 16 byte alignment in another. 
Alignment is important for determining the number of 32 byte chunks a loop is spanning.

\begin{lstlisting}[language=C, escapeinside=``, caption={Example of bias from link order. Instruction alignment of the loop changes with ordering. }]
`loop.c`
#define B22(n) b += i*i*a + 1
volatile int i = 42;

int loop(int a) {
    register int b = i == 42 ? 1 : 0;
    do {
        // 10 bytes mov(6), test(2), jne(2)
        if (!i) B22(1); 
        if (!i) B22(2);
        if (!i) B22(3);
        if (!i) B22(4);
        if (!i) B22(5);
        if (!i) B22(6);
        if (!i) B22(7);
        if (!i) B22(8);
        if (!i) B22(9);
        if (!i) B22(10);
        if (!i) B22(11);
        // 27 bytes for increment and test
    } while (i++ < 0x12345678);
    return b;
}
`foo.c`
int foo(int n) {
    int a = 1;
    return a;
}
`main.c`
#include <stdio.h>
int main() {
    int i = loop(10);
    int f = foo(10);
    printf("Loop: %d\n", i);
    printf("Foo: %d\n", f);
    return 0;
}
\end{lstlisting}

\subsection{Other Triggers of Bias Effects}
The focus of previous work, and elaborated in this section, is measurement bias caused by altering link ordering.
The important factor is not the process of linking itself, but the effect it has on final code layout in the compiled binary.

\paragraph{Order of functions within a file}
As with link order, the order of functions within a source file usually determines their relative position in the text segment of the compiled binary. 
Given a program with functions foo and bar, listing foo before bar and vice versa generates two “different” programs with respect to memory layout. 
GCC prints functions to the text segment in the order they appear in the source code.

\paragraph{Length of external symbols}
Symbols such as function names are placed in a symbol table section in the ELF file.
Symbols that needs to be resolved dynamically are allocable, and mapped to virtual address space before the text segment. 
This means that names of external symbols resolved at run-time does affect code layout. 
Longer function names (or more external dependencies) will offset the text segment because the size of allocable segments related to dynamic linking changes. 
his includes the “dynsym”, “dynstr”, and “rela” sections in an executable made with GCC.

Some particularly devious “bugs” can occur from this. 
Consider a scenario where a program does not align properly to utilize the Loop Stream Detector. 
Assume a printf statement is added to a completely unrelated part of the code. 
The additional external symbol offsets instruction alignment such that the LSD can be used, giving a huge speedup.
Later removing the printf statement will reduce the size of external dependencies, and again result in worse performance.

Any interaction that affects code layout can potentially trigger performance cliffs caused by the Loop Stream Detector.

@@

The Loop Stream Detector can provide significant speedup, but subtle changes to code layout sometimes prevents this optimization to be utilized. 
Programmers rarely care, or even know, about intricate details such as the addresses and alignment of code. 
We expect, and probably rightfully so, that sensible compilers and linkers will organize code somewhat optimally. 
In cases where this is not true, weird effects that might be categorized as “bias” occurs. 
For developers, it is useful to be aware of the fact that any change to instruction addresses can potentially have a huge impact on performance.

>>>>

\documentclass[a4paper,10pt,twocolumn,twoside]{article}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{filecontents}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{tikz}
\usepackage{listings}
\usepackage{authblk}
\usepackage{caption}

@@

\subsection{Analysis}

@@

\subsection{Avoid Aliasing -- A Proof of Concept}
Addresses of automatic variables can not be determined statically, because the position of stack at runtime is generally unknown. 
In addition to being offset by environment variables, the stack address can also be perturbed by other factors such as address layout randomization. 
Although one can not easily know if a collision is going to happen for a given environment, one can try to change the program to account for possible alias effects.
The following strategy is a proof of concept of how alias-free code can be generated in this particular case.
If the addresses do alias, branch to an alternative but semantically equivalent code path.
Allocate a new set of variables to avoid aliasing.

\begin{lstlisting}[float=h, language=C, caption={Dynamically detect aliasing case, and avoid by pushing another stack frame.}, label={lst:loopkernel}, frame=lines]
static int i, j, k;

int main() {
    int g = 0, inc = 1;

    if ((((long)&inc) & 0xfff) == (((long)&i) & 0xfff) || 
        (((long)&g) & 0xfff) == (((long)&i) & 0xfff))
        return main();

    for (; g < 65536*10; g++) {
        i += inc;
        j += inc;
        k += inc;
    }

    return 0;
}
\end{lstlisting}

@@

\begin{lstlisting}[float=t, language=C, caption={Naive implementation of convolution. Highly sensitive to aliasing between input and output arrays.}, label={lst:conv}, frame=lines]
static float k[5]={0.1,0.25,0.3,0.25,0.1};
void conv(int n, float *in, float *out) {
    int i, j;
    for (i = 2; i < n-2; ++i) {
      out[i] = 0;
      for (j = 0; j < 5; ++j)
          out[i] += in[i-2+j] * k[j];
    }
}
\end{lstlisting}

@@

@@
